---
title: "Pivoting, Combining Datasets, and Tables in R"
author: "Nina Galanter"
date: "\today"
output:
  pdf_document:
    toc: no
  html_document:
    toc: no
subtitle: BIOST 311, Discussion Section Week 4
---

```{r setup, include=FALSE}
library(readr)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```


# Introduction

Today we are going to learn how to pivot datasets and combine them, which can be an important first step before conducting a regression analysis. We will also look at creating summary tables with `kable` and `summarize`.

# Pivoting

In order to run regression, our datasets must have one observation in each row, and every variable should be a column. Data often doesn't look like this initially, which is where pivoting comes in.

## Pivot_wider

Sometimes we will find datasets where different variables are in different rows of the data, instead of in different columns. One example of this is the `us_rent_income` dataset in the `tidyr` package.

```{r viewing rent data}
library(tidyverse)


data("us_rent_income")

head(us_rent_income)


```

\newpage

Instead of there being a row for each geographical area which includes a variable for rent and a variable for income, this dataset has rent and income as separate row. We can use the `pivot_wider` function to fix this. 

```{r wider example}
us_rent_income2 <- us_rent_income %>% pivot_wider( names_from = variable,
                                                   values_from=c(estimate,moe))

head(us_rent_income2)
```


Now we have a more reasonable dataset - we have a separate variables for income and rent. The `names_from` argument of `pivot_wider` tells us from which column(s) we should get the names of the new variables we are adding. The `values_from` column(s) tell us the values of these new variables.

**Exercise 1**

```{r wider data}

#creating synthetic blood pressure data

blood_pressure_data <- data.frame(id = c(1,1,2,2,3,3,4,4,5,5,6,6,7,7,8,8,9,9,10,10),
                                  Type = rep(c("Systolic","Diastolic"),10),
                                  Value = c(rnorm(n = 20, mean = 80, sd = 5)))

head(blood_pressure_data)

```


For the first exercise, use the `blood_pressure_data` dataset created above, Currently, this dataset has systolic and diastolic blood pressure measurements in different rows. Use `pivot_wider` to make the two types of measurements different variables.

```{r wider exercise}

```

\newpage

## Pivot_longer

Often, we will have the opposite problem with our data, where there are multiple observations per row. We can fix this with `pivot_longer`. Below is an example using the `billboard` dataset in tidyr. We have data for each song that was in the Billboard top 100 (a music popularity ranking) in 2000. There is a variable for the rank of the song each week. However, if we want to plot the rating of songs over time, we need each obesrvation to be a specific song during a specific week. 

```{r longer example}
data(billboard)

head(billboard)

billboard2 <- billboard %>%
  pivot_longer(
    cols = 4:79,
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  )

head(billboard2)


```

This time, we took all of our week columns and turned them into a week variable and a rank variable. The `cols` argument specified which columns to use. The `names_to` argument determined the name of variable for the column names. the `values_to` argument determined the name of the variable for the column values. The `names_prefix` argument tells us to get rid of the prefix "wk" when making the names variable. Finally, as songs are usually not in the top 100 for the entire year, we use the `values_drop_na` variable to get rid of the weeks that each song wasn't in the top 100.

**Exercise 2**

Take the dataset `children` below (from the`household` dataset in the `tidyr` package) and use `pivot_longer` to make a new version where each child is a different observation (row).

```{r longer exercise}
data("household")

children <- household %>% select(family,name_child1,name_child2)


```

# Merging Datasets

When dealing with complicated data, we often end up with several different datasets. For example, clinical trial data will consist of spreadsheets recording different types of outcomes and spreadsheets of adverse events. In order to analyze data, however, we need all the variables in our analysis to be in the same data frame in R. We can use the join functions to do this. 

```{r join example}
data1 <- data.frame(Subject = c(2,1,3), treatment = c(0,1,1))

data2 <- data.frame(Subject = c(2,3,4), outcome = c(1,1,0))

```

Here we've created two datasets of subjects. We want to combine them into one dataset to look at the relationship between treatment and outcome. However, `data1` is missing subject 4 and `data2` is missing subject 1. Each of the join functions handles this differently.

```{r joins}
inner_join(data1,data2)

left_join(data1,data2)

right_join(data1,data2)

full_join(data1,data2)
```

**Exercise 3**

Describe how each of these different joins handle the situation where some subjects aren't shared between the datasets.

# Summary Tables

We can use the `summarize` (you can also spell it `summarise`), `group_by`, and `kable` functions to create nice summary tables. `kable` is part of the `knitr` package.

For this example, say that I want to get a summary of the data comparing the groups who did and did not participate in the first steps program.

```{r tables example}
library(knitr)

births <- read_csv(url("https://raw.githubusercontent.com/cwolock/BIOST311/main/Datasets/births.csv"))

births %>% mutate(firstep2 = ifelse(firstep == 1,"Yes","No")) %>%
  group_by(firstep2) %>% 
  summarize(
    Count = n(),
    Age = paste0(round(mean(age),2)," (",round(sd(age),2),")"),
    "Birth Weight" = paste0(round(mean(bwt),2)," (", round(sd(bwt),2) ,")"),
    Smoke = paste0(sum(smoke)," (", round(mean(smoke) ,2)*100,"%)"),
    Married = paste0(sum(married)," (", round(mean(married) ,2)*100,"%)")) %>% 
  rename("First Step" = firstep2) %>%
  kable(caption = "Variable summaries by first step participation. The format is mean (sd) for quantitative variables and count (%) for binary variables.")


```

The example above uses the `round` and the `paste0` function. The `paste0` function just pastes together all of its arguments into one string. We can also include a caption for our table by using the `caption` argument of the `kable` function.

**Exercise 4**

Again using the `births` dataset, get the same table as before, except now comparing the subjects who did and did not drink alcohol. Also change the table from mean and sd to median and IQR.

```{r tables exercise}

```

