---
title: "Inline R code"
author: "Nina Galanter"
date: '2023-04-11'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r preliminaries, message = FALSE, results = "hide"}
library(tidyverse)
# install.packages("car")
library(car)
data(UN)
# ?UN # run this line for details on the dataframe
```

## Inline Code

When interpretting models in R markdown, it may be helpful to use inline code in R, where we put R code in written parts  of an R markdown. Two examples are below.


```{r example1, message = FALSE, results = "hide"}

mod <- lm(data = UN, lifeExpF ~ pctUrban)

```

We estimate that, comparing two groups of countries with a difference of 1 percentage point urban, female life expectancy is `r round(coef(mod)[2],2)` years higher for the more urban countries.  Based on a 95% confidence interval, this estimated difference would not be judged unusual if the true difference were between `r round(confint(mod)[2,1],2)` and `r round(confint(mod)[2,2],2)` years.

```{r example2, message = FALSE, results = "hide"}

mod <- lm(data = UN, lifeExpF ~ pctUrban)

beta1 <- coef(mod)["pctUrban"]

ci_low <- confint(mod)["pctUrban", "2.5 %"]
ci_high <- confint(mod)["pctUrban", "97.5 %"]

```

We estimate that, comparing two groups of countries with a difference of 1 percentage point urban, female life expectancy is `r round(beta1,2)` years higher for the more urban countries. Based on a 95% confidence interval, this estimated difference would not be judged unusual if the true difference were between `r round(ci_low,2)` and `r round(ci_high,2)` years.